Management and effectiveness of community-based marine reserves in small-scale fisheries

### 2018-06-08 docs/paper/Villasenor-Derbez.tex
Made changes to manuscript body, just missing details on discussion

\title[Marine reserve effectiveness]{Enabling conditions for effective community-based marine reserves in
small-scale fisheries}

@@

Coastal marine ecosystems provide livelihoods for small-scale fishers
and coastal communities around the world. However, overfishing and
unsustainable fishing practices threaten the marine environment and
jeopardize the wellbeing of coastal communities. A common approach to
protect the environment and recover overexploited stocks is to implement
no-take marine reserves (areas where all extractive activities are
off-limits). In small-scale fisheries, these are sometimes implemented
as community-based reserves, where a group of fishers collectively agree
to close an area to fishing. While we know that reductions in fishing
effort are followed by a series of ecological benefits (increased
biomass, abundance, and species diversity), we do not fully understand
how environmental and governance dynamics influence the conservation and
fisheries benefits of community-based marine reserves. In this work, we
evaluate the ecological outcomes of four reserves established by three
coastal communities in temperate and tropical ecosystems of Mexico. By
combining causal inference techniques with an operationalization of the
social-ecological systems framework, we identify the environmental and
social conditions that enable reserve effectiveness. Our results show a
strong interaction between environmental variation and community
organization, which influences reserve effectiveness. For example, the
most effective reserve had strong governance structures accompanied with
low environmental variability. Thus, even when a community is well
organized (and reserves are well enforced), environmental variation can
hinder the benefits of a reserve, and vice versa. Our results are
particularly relevant under present changing climate conditions, as they
can better inform management and decision making.

@@

\textbf{Last\_update}: 2018-05-13

@@

Marine ecosystems around the world sustain significant impacts due to
overfishing and unsustainable fishing practices
\citep{halpern_2008-dK,worm_2006-IB,pauly_2005-qV}. A common approach to
manage the spatial distribution of fishing effort and recover stocks is
through the implementation of marine reserves (\emph{i.e.} areas where
all fishing activities are off--limits; MRs)
\citep{afflerbach_2014-HP,krueck_2017-J1,sala_2017-69}.

Marine reserve science has largely focused on understanding the
ecological effects of these areas, which include increased biomass,
richness, and densities of organisms within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}, climate change
mitigation \citep{roberts_2017-J9}, and protection from environmental
variability \citep{micheli_2012-EU}. However, there is considerably less
literature focusing on the relationship between socioeconomic and
governance structures and their relationship to ecological effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_} or benefits to
fisheries \citep{krueck_2017-J1}; evaluations of marine reserves rarely
provide a holistic view of the social-ecological system
\citep{lpezangarita_2014}.

While ecological factors (\emph{e.g.} habitat representation, initial
state of protection, connectivity to other protected areas) can
determine the success of a MR, their effectiveness also depends on the
socioeconomic and governance settings under which they are implemented
and managed. Literature shows that many non-ecological characteristics
can play an equally important role in the effectiveness of MRs. For
example, age of a reserve (\emph{i.e.} time since its implementation)
and size were key to the effectiveness of MRs in Palau
\citep{friedlander_2017-oI}. In the Mediterranean,
\citet{difranco_2016-Xw} identify that surveillance and enforcement,
presence of a management plan, and involvement of fishers in management
and decision--making along with promotion of sustainable fishing
practices were the key factors that increased stock health and income to
fishers. At a global level, \citet{edgar_2014-UO} indicate that
enforcement, age, size, and isolation were important factors determining
effectiveness of the reserves.

Marine Reserves in Mexico have been commonly implemented as ``core
zones'' within Biosphere Reserves that are administered by the National
Commission of Protected Areas (\emph{Comisión Nacional de Áreas Marinas
Protegidas}, CONANP). While CONANP has made efforts to have a
participatory process, the implementation of these zones is still
characterized by top--down approaches. This motivated Civil Society
Organizations (CSOs) to work with coastal communities to implement
community--based marine reserves \citep{uribe_2010-u2}, which are
usually established within a TURFs; thus making them TURF--reserves
\citep{afflerbach_2014-HP}. This bottom--up approach allows fishers to
design their own reserves, which increases compliance and
self--enforcement
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8}. However,
these reserves still lack legal recognition, making them vulnerable to
poaching. In 2014, a new norm \citep{nom} allowed fishers to request the
legal recognition of a community--based reserve under the name of
``Fishing Refugia'' (\emph{Zona de Refugio Pesquero}, FR). Since then,
\textbf{39} FR have been implemented along the Pacific, Gulf of
California, and Mexican Caribbean coastlines, but no formal evaluation
of their effectiveness has taken place.

We combine causal inference techniques \citep{depalma_2018} and the
social-ecological systems framework \citep{ostrom_2009-hg} to provide a
comprehensive ecological and socioeconomic evaluation of community-based
marine reserves in three coastal communities in Mexico. The objective of
this work is twofold: i) Provide the first evaluation of
community--based marine reserves in Mexico, and ii) provide a
comprehensive evaluation of the social--ecological system to identify
how socioeconomic and governance characteristics relate to ecological
effectiveness.

@@

We focus our evaluation in three coastal communities from the Pacific
coast of Baja California and the Mesoamerican Reef System (Fig
\ref{fig:map}). Fishing activities in all communities are assambled
around fishing cooperatives that hold Territorial Use Rights for
Fisheries (TURFs). Isla Natividad (IN) lies west of the Baja California
Peninsula (Fig \ref{fig:map}B), where kelp forests (\emph{Macrocystis
pyrifera}) and rocky reefs are the predominant habitats. The island is
home to a the \emph{Sociedad Cooperative de Producción Pesquera (SCPP)
Buzos y Pescadores de la Baja California}, spiny lobster
(\emph{Panulirus interruptus}) represents an important fishery. However,
other resources like finfish (yellow-tail jack, \emph{Seriola lalandi}),
sea cucumber (\emph{Parastichopus parvimensis}), red sea urchin
(\emph{Mesocentrotus franciscanus}), snail (\emph{Megastraea turbanica}
y \emph{M. undosa}), and abalone (\emph{Haliotis spp}, until 2010) are
also important sources of income. In 2006, the community decided to
implement two community--based marine reserves within their fishing
grounds, seeking to recover depleted stocks of invertebrate species
(mainly lobster and abalone). Until today, these reserves are yet to be
legally recognized as Fishing Refugia but count with full support from
the community.

The other two communities are Maria Elena (ME) and Punta Herrero (PH;
Fig \ref{fig:map}C) in the Yucatan Peninsula, where coral reefs and
mangroves are the representative coastal ecosystems. ME is a fishing
camp --visited intermittently during the fishing season-- belonging to
the Cozumel fishing cooperative (\emph{SCPP Cozumel}). PH is also home
to a fishing cooperative (\emph{SCPP José María Azcorra}). The main
source of income to both communities is the Caribbean spiny lobster
(\emph{Panulirus argus}). These communities also target finfish in the
off season, mainly snappers (Lutjanidae) and groupers (Serranidae). ME
established eight marine reserves in 2012, and PH established four
marine reserves in 2013 and an additional community-based (\emph{i.e.}
not legally recognized). All these reserves are legally recognized as
Fishing Refugia.

@@

We use three main sources of information to evaluate these reserves.
Ecological data come from the annual ecological monitoring of reserve
and control areas, carried out by members from each community and
personnel from the Mexican CSO ``Comunidad y Biodiversidad''
(\href{www.cobi.org.mx}{COBI}). These monitorings record richness and
abundances of fish and invertebrate species in the reserves and control
sites. Size structures are also collected in fish census. We define
control sites as regions with habitat characteristics similar to the
corresponding reserves, and that presumably had the same probability of
being selected as reserves during the design phase. From all the
reserves in these three communities, we use the ones that have data for
reserve and control sites before and after the implementation of the
reserve. This provides us with a Before-After-Control-Impact
(\emph{i.e.} BACI) sampling design that allows us to capture and control
for temporal and spatial dynamics \citep{depalma_2018,ferraro_2006-oW}.
BACI designs and causal inference techniques have proven effective to
evaluate marine reserves, as they allow us to causally attribute
observed changes to the intervention
\citep{moland_2013-VP,Villasenor-Derbez_2018}. All reserves were
surveyed annually from at least one year before implementation until
2016. Table \ref{table:com_sum} shows a summary of sampling effort and
year of implementation of the analyzed reserves.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Location of the three
coastal communities studied (A). Isla Natividad (B) is located off the
Baja California Peninsula, Maria Elena (C) and Punta Herrero (D) are
located in the Yucatan Peninsula.}
\end{figure}

@@

Socioeconomic data come from landing receipts reported to the National
Commission for Aquaculture and Fisheries (\emph{Comisión Nacional de
Acuacultura y Pesca}; CONAPESCA). Data contain monthly lobster landings
(Kg) and value (MXP) from 2000 to 2014. This information was aggregated
by year, and economic values were adjusted by the Consumer Price Index
\citep{oecd_2017-VV} via Eq \ref{eqn:cpi}.

@@

We use a set of governance indicators to analyze the governance
structures of each cooperative \citep{leslie_2015-na}. The indicators
(Table \ref{table:gov_ind}) focus on the resource system (four
indicators), governance system (seven indicators), resource units (three
indicators) and actors (three indicators). These were collected at the
community--level from official documents used in the creation and
designation of the marine reserves
\citep{dof_website_2012,dof_website_2012} and based on the authors'
experience and knowledge of the communities.

@@

We evaluate the effect thar marine reserves have had on four biological
and two socioeconomic indicators (Table \ref{table:indicators}). Recall
that reserves were implemented to protect lobster and other benthic
invertebrates. However, we also contemplate fish species that are also
targeted in the off-season and might receive collateral protection from
the reserves.

@@

\caption{\label{tab:unnamed-chunk-3}\label{table:indicators}List of indicators used to evaluate the effectiveness of marine reserves, grouped by category.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Category & Indicador & Units\\
\hline
Biological & Lobster density & org / $m^2$\\
\hline
Biological & Invertebrate density & org / $m^2$\\
\hline
Biological & Fish biomass & org / $m^2$\\
\hline
Biological & Fish density & org / $m^2$\\
\hline
Socioeconomic & Income from target species & M MXP\\
\hline
Socioeconomic & Landings from target species & Metric Tones\\
\hline
\end{tabular}
\end{table}

\caption{\label{tab:unnamed-chunk-4}\label{table:gov_ind}Indicators used for the operationalization of the SES framework \citep{leslie_2015-na}}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}|>{\raggedright\arraybackslash}p{8em}|>{\raggedright\arraybackslash}p{8em}|>{\raggedright\arraybackslash}p{8em}}
\hline
Indicator & Isla Natividad & Maria Elena & Punta Herrero\\
\hline
\textbf{Resource systems (RS)} & \textbf{} & \textbf{}\\
\hline
TURF presence & Yes & Yes & Yes\\
\hline
Type of ecosystem & Kelp Forest / Rocky Reefs & Coral Reef & Coral Reef\\
\hline
Intensity of environmental Disturbance & El nino event & Hurricanes & Hurricanes\\
\hline
Location & Island & Coastal & Coastal\\
\hline
\textbf{Governance systems (GS)} & \textbf{} & \textbf{}\\
\hline
Fishing cooperative & Yes & Yes & Yes\\
\hline
Involved actors & COBI, Stanford, REBIVI & Alianza Kanan Kay, COBI, CONANP, Coop, CONAPESCA, Oceanus, FCyRH, FHMM, & Alianza Kanan Kay, COBI, CONANP, Coop, CONAPESCA, Oceanus, FCyRH, FHMM,\\
\hline
Presence of an inter-cooperative structure & Fedecoop & Non & Non\\
\hline
Fishing Regulations & Size limits, seasonal closures, quotas & Size limits, seasonal closures & Size limits, seasonal closures\\
\hline
Enforcement technology & Boats & Boats & Land enforcement\\
\hline
MR enforcement &  &  & \\
\hline
Cooperative regulations &  &  & \\
\hline
\textbf{Resource Units (RU)} & \textbf{} & \textbf{}\\
\hline
Adult targeted species mobility & 1km & 30km & 30km\\
\hline
Targeted species longevity (years) &  &  & \\
\hline
Price of targeted species &  &  & \\
\hline
\textbf{Actors (A)} & \textbf{} & \textbf{}\\
\hline
Leadership &  &  & \\
\hline
Level of illegal fishing & 1 & 1 & 3\\
\hline
Presence of alternative livelihoods &  &  & \\
\hline
\end{tabular}
\end{table}

Biological indicators are analyzed with a difference--in--differences
analysis (Eq \ref{eqn:reg_bio}), which allows us to estimate the effect
that the reserve has on the biological indicators by comparing trends
across time and treatments
\citep{moland_2013-VP,Villasenor-Derbez_2018}. The analysis is performed
with a multiple linear regression of the form:

@@

Where year--fixed effects are represented by \(\gamma_{it} Year_t\), and
\(\beta Zone_i\) captures the difference between reserve (\(Zone = 1\))
and control (\(Zone = 0\)) sites. The interaction term
\(\lambda_{it} Year_t\times Zone_i\) represents represent the mean
change in the indicator inside the reserve, for year \(t\), with respect
to the first year of evaluation in the control site (See Table
\ref{table:com_sum}). When evaluating biomass and abundances, we include
species--fixed effects (\(\sigma_j\)).

@@

\begin{equation}
I_i = \alpha + \gamma_{t} Year_t + \beta Treated_i + \lambda_{t} Year_t\times Treated_i + \sigma_jComm_j +\epsilon
\label{eqn:soc_reg}
\end{equation}

The model interpretation remains as for Eq \ref{eqn:reg_bio}, but in
this case the \(Treated\) dummy variable indicates if the community has
a reserve (\(Treated = 1\)) or not (\(Treated = 0\)) and
\(\sigma_jComm\) captures community-level fixed-effects. This approach
allows for a causal attribution of the observed changes to the reserve.
All model coefficients were estimated via ordinary least--squares and
heteroskedastic--robust standard errors \citep{zeileis_2004-7n}.

These regressions allows us to make a causal link between the
implementation of marine reserves and the observed trends by accounting
for temporal and spatial dynamics \citep{depalma_2018}. The effect of
the reserve is captured by the \(\lambda_t\) coefficient, and represents
the difference observed between the control site before the
implementation of the reserve and the reserve site at time \(t\) after
controlling for other time and space variations (\emph{i.e.}
\(\gamma_t\) and \(\beta\) respectively).

@@

The following sections present the effect that marine reserves had on
each of the biological and socioeconomic indicators for each coastal
community. Results are presented in terms of the difference through time
and across sites, relative to the control site on the year of
implementation (\emph{i.e.} effect size). We also provide an overview of
the governance settings of each community, and discuss how these
dimensions might be related to the effectiveness of the reserves.

@@

All indicators show idiosyncratic responses through time for each
community. Figure \ref{fig:indicators}A shows how lobster densities
increase inside the reserve for Isla Natividad an Punta Herrero within
the first years, but the effect is aroded through time. However, these
effects are in the order of 0.2 extra organisms \(m^{-2}\) and are not
significantly different from zero (\(p > 0.05\)). The only case when
lobster densities showed a significant positive effect (\(p < 0.1\)) was
for Isla Natividad on the sixth year (\emph{i.e.} 2012), a year after
hypoxia events described in \citet{micheli_2012-EU} caused mass
mortality of organisms. Biomass was only evaluated for fish data (Fig
\ref{fig:indicators}B), and no changes were observed (\(p > 0.1\)),
Punta Herrero even showed negative insignificant trends in the order of
0.01. Invertebrate and fish densities (\ref{fig:indicators}C-D) show
similar patterns of lack of effectiveness with effect sizes oscillating
around zero. Full tables with model coefficients are presented in the
supplementary materials (\textbf{S1 Table}, \textbf{S2 Table},
\textbf{S3 Table}).

@@

Lobster landings and revenue were only available for Isla Natividad and
Maria Elena (Fig \ref{fig:lobsters}). For all years befor
implementation, the effect sizes are close to zero, indicating that the
control and treatment sites track each other well. However, effect sizes
do not change after the implementation of the reserve. Again, the
negative coeficient observed for Isla Natividad on year 5 correspond to
the 2011 hypoxia events. The only positive change observed in lobster
landings is for Isla Natividad in 2014 (\(p < 0.1\)). The trhee years of
post-implementation data for Maria Elena do not show a significant
effect of the reserve. Isla Natividad shows higher revenues after the
iumplementation of the reserve, as compared to the control communities.
However, these changes are also not ignificantly different and present
an increased variation. All regression coefficients for each community
and indicator are presented in \textbf{S4 Table}.

@@

Although we have little information on the social dimension of these
fisheries, using the SES framework indicators (Table
\ref{table:gov_ind}), we can analyze the performance of each governance
system (Table \ref{table:gov_res}).

@@

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{\label{fig:unnamed-chunk-5}\label{fig:indicators}Effect sizes for
marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME;
blue triangles), and Punta Herrero (PH; green squares) for lobster
densities (\emph{Panulirus spp}; A), fish biomass (B), invertebrate
densities (C), and fish densities (D). Plots are ordered by survey type
(left column: invertebrates; right column: fish). Points are jittered
hotizontally to avoid overplotting. Points indicate the effect size, and
errorbars standard errors. Years have been centered to year of
implementation.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-7-1.pdf}
\caption{\label{fig:unnamed-chunk-7}\label{fig:lobsters}Effect sizes for
lobster catches (A) and revenues (B) in at Isla Natividad (IN; red
circles) and Maria Elena (ME; blue triangles)}
\end{figure}

\caption{\label{tab:unnamed-chunk-8}\label{table:gov_res}Analysis of the fishing cooperatives based on the Social-Ecological systems framework \citep{mcginnis_2014}.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}|>{\raggedright\arraybackslash}p{10em}|l|l|l}
\hline
 & Indicator & Isla Natividad & Maria Elena & Punta Herrero\\
\hline
\textbf{Resource systems (RS)} & \textbf{} & \textbf{} & \textbf{}\\
\hline
RS2 – Clarity of system boundaries & TURF presence & High & High & High\\
\hline
RS3 – Size of resource system &  &  &  & \\
\hline
RS5 – Productivity of system & Type of ecosystem & High & High & High\\
\hline
RS7 – Predictability of system dynamics & Intensity of environmental disturbance & Low (ENSO) & High & High\\
\hline
RS9 – Location & Proximity to other communities/cities & Isolated & Not Isolated & Not Isolated\\
\hline
\textbf{Governance systems (GS)} & \textbf{} & \textbf{} & \textbf{}\\
\hline
GS1 – Government organizations & Presence of fishing cooperatives & Yes & Yes & Yes\\
\hline
GS2 – Nongovernment organizations & Involved actors & Yes & Yes & Yes\\
\hline
GS3 – Network structure & Presence of an inter-cooperative structure & Yes & No & No\\
\hline
GS4 – Property-rights systems & TURF presence & Yes & Yes & Yes\\
\hline
GS5 – Operational-choice rules & Fishing Regulations / MPA enforcement / Enforcement technolofy & Yes & Yes & Yes\\
\hline
GS6 – Collective-choice rules & Cooperative regulations & Yes & Yes & Yes\\
\hline
GS7 – Constitutional-choice rules &  &  &  & \\
\hline
\textbf{Resource units (RU)} & \textbf{} & \textbf{} & \textbf{}\\
\hline
RU1 – Resource unit mobility & Targeted species home range & Low & Medium & Medium\\
\hline
RU2 – Growth or replacement rate & Max age of targeted species & Low & Medium & Medium\\
\hline
RU4 – Economic value & Price of targeted species & high & High & high\\
\hline
\textbf{Actors (A)} & \textbf{} & \textbf{} & \textbf{}\\
\hline
A1 – Number of relevant actors &  & 98 &  & \\
\hline
A2 – Socioeconomic attributes &  &  &  & \\
\hline
A5 – Leadership/entrepreneurship & Leadership & High & High & High\\
\hline
A6 – Norms (trust-reciprocity)/social capital--- (Based on illegal fishing) & Level of illegal fishing & High & High & Low\\
\hline
A8 – Importance of resource (dependence) & Presence of alternative livelihoods & High & High & High\\
\hline
\end{tabular}
\end{table}

@@

@@

>>>>

\title[Community-based marine reserves]{Management and effectiveness of community-based marine reserves in
small-scale fisheries}

@@

Al finalizar revisiones

@@

\textbf{Last update}: 2018-06-08

@@

Marine ecosystems around the world sustain significant impacts due to
overfishing and unsustainable fishing practices
\citep{halpern_2008-dK,worm_2006-IB,pauly_2005-qV}. A common approach to
manage the spatial distribution of fishing effort to recover stocks and
preserve biodiversity is through the implementation of marine reserves
(MRs). These areas allow bounded populations to recover by limiting all
extractive activities \citep{halpern_2002}. The science of MRs has
largely focused on understanding the ecological effects of these areas,
which include increased biomass, richness, and densities of organisms
within the protected regions, climate change mitigation, and protection
from environmental variability
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69,roberts_2017-J9,micheli_2012-EU}.
Modelling studies show that the fishery benefits of marine reserves
depend on initial stock status and the type of management under which
the fishery operates \citep{hilborn_2006}, as well as reserve size and
the amount of larvae exported from these \citep{krueck_2017-J1}. Other
research has focused on the relationship between socioeconomic and
governance structures and their relationship to ecological effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_}. However, few
studies simultaneously evaluate MRs from all these perspectives
(\emph{e.g.} \citet{lpezangarita_2014}).

While ecological factors like habitat representation within the MR or
connectivity to other areas can determine the success of a MR, its
effectiveness also depends on the socioeconomic and governance settings
under which they are implemented and managed. Literature shows that many
non-ecological characteristics can play an equally important role in the
effectiveness of MRs. For example, age of a reserve (\emph{i.e.} time
since its implementation) and size were key to the effectiveness of MRs
in Palau \citep{friedlander_2017-oI}. In the Mediterranean,
\citet{difranco_2016-Xw} identify that surveillance and enforcement,
presence of a management plan, and involvement of fishers in management
and decision--making along with promotion of sustainable fishing
practices were the key factors that increased stock health and income to
fishers. At a global level, enforcement, age, size, and isolation are
important factors that determine the effectiveness of the reserves
\citep{edgar_2014-UO}. However, MRs represent part of a more complex
social-ecological systems governed by the interaction between humans and
the environment. In this work we evaluate community-based marine
reserves in Mexico using a triple bottom line approach by using
ecological, socioeconomic, and governance indicators that provide a
holistic view of the reserves \citep{halpern_2013}.

There are three main approaches to implement MRs, each of which have its
pros and cons. We describe them in the context of Mexican MRs, but argue
that at least the first two apply elsewhere. MRs were historically
implemented and managed by a government agency, in this case the
National Commission of Protected Areas (\emph{Comisión Nacional de Áreas
Marinas Protegidas}, CONANP). While CONANP has made efforts to have a
participatory process, the implementation of these zones is still mainly
a top-down process. A second approach is the implementation of
community-based marine reserves within areas of exclusive access
(\emph{i.e.} TURFs), thus making them TURF-reserves
\citep{afflerbach_2014-HP}. Community-based spatial closures ocurr in
other places, like the \emph{kapu} or \emph{ra'ui} areas in the Pacific
Islands \citep{bohnsack_2004,johannes_2002}. This bottom-up approach
increases compliance and self-enforcement
\citep{gelcich_2015-Gw,espinosaromero_2014-PY,beger_2004-Y8}. However,
these lack legal recognition and rely on the exclusive access granted by
the TURF, making it impossible to enforce without one. In an effort to
provide a legal framework for these reserves, Civil Society
Organizations (CSOs) served as the bridge between fishers and
government. In 2014, a new norm \citep{nom} allowed fishers to request
the legal recognition of a community-based reserve under the name of
``Fishing Refugia'' (\emph{Zona de Refugio Pesquero}). These can be
implemented as temporal or partial eservs, which can protect one, some,
or all resources within them. Since then, \textbf{39} of these have been
implemented along the Pacific, Gulf of California, and Mexican Caribbean
coastlines, but no formal evaluation of their effectiveness has taken
place.

This work combines causal inference techniques and the social-ecological
systems framework to provide a holistic evaluation of community-based
marine reserves in three coastal communities in Mexico. The objective of
this work is twofold. First, provide a triple bottom line evaluation of
the effectiveness of community-based marine reserves that can inform
similar processes in other countries. And second, perform the first
formal evaluation of Fishing Refugia in Mexico and identify areas where
improvement or adjustment might result in increased effectiveness. On
both cases, we draw from the lessons learned and provide management
recommendations to ensure or improve the effectiveness of
community-based marine reserves.

@@

We evaluate community-based marine reserves from three coastal
communities located in the Pacific coast of Baja California and the
Mexican Caribbean (Fig \ref{fig:map}). All communities are organized as
fishing cooperatives that hold Territorial Use Rights for Fisheries
(TURFs). Isla Natividad (IN) lies west of the Baja California Peninsula
(Fig \ref{fig:map}B), where kelp forests and rocky reefs are the
predominant habitats. The island is home to the \emph{Sociedad
Cooperativa de Producción Pesquera (SCPP) Buzos y Pescadores de la Baja
California}, whose main resource by value is the spiny lobster
(\emph{Panulirus interruptus}). However, other resources like finfish
(yellow-tail jack, \emph{Seriola lalandi}), sea cucumber
(\emph{Parastichopus parvimensis}), red sea urchin (\emph{Mesocentrotus
franciscanus}), snail (\emph{Megastraea turbanica} y \emph{M. undosa}),
and abalone (\emph{Haliotis spp}) are also important sources of income.
In 2006, the community decided to implement two community-based marine
reserves within their fishing grounds to protect commercially important
invertebrate species (mainly lobster and abalone). Until today, these
reserves are yet to be legally recognized as Fishing Refugia but count
with full support from the community.

The other two communities are Maria Elena (ME) and Punta Herrero (PH;
Fig \ref{fig:map}C) in the Yucatan Peninsula, where coral reefs and
mangroves are the representative coastal ecosystems. ME is a fishing
camp --visited intermitently during the fishing season-- belonging to
the Cozumel fishing cooperative (\emph{SCPP Cozumel}); PH is home to the
\emph{SCPP José María Azcorra} cooperative. Their main fishery is the
Caribbean spiny lobster (\emph{Panulirus argus}), but they also target
finfish in the off season, mainly snappers (Lutjanidae) and groupers
(Serranidae). ME established eight marine reserves in 2012, and PH
established four marine reserves in 2013 and an additional
community-based (\emph{i.e.} not legally recognized). All these reserves
are legally recognized as Fishing Refugia.

@@

We use three main sources of information to evaluate these reserves.
Ecological data come from the annual ecological monitoring of reserve
and control areas, carried out by members from each community and
personnel from the Mexican CSO \emph{Comunidad y Biodiversidad}
(\href{www.cobi.org.mx}{COBI}). Trained divers record richness and
abundances of fish and invertebrate species in the reserves and control
sites. Size structures are also collected for fish evaluations. We
define control sites as regions with habitat characteristics similar to
the corresponding reserves, and that presumably had a similar
probability of being selected as reserves during the design phase. We
focus our evaluation for sites where data are avaialable for reserve and
control sites, before and after the implementation of the reserve. This
provides us with a Before-After-Control-Impact (\emph{i.e.} BACI)
sampling design that allows us to capture and control for temporal and
spatial dynamics \citep{depalma_2018,ferraro_2006-oW}. BACI designs and
causal inference techniques have proven effective to evaluate marine
reserves, as they allow us to causally attribute observed changes to the
intervention \citep{moland_2013-VP,Villasenor-Derbez_2018}. All sites
were surveyed annually, and at least once before implementation of the
reserves. Table \ref{table:com_sum} shows a summary of the reserves
included in this study.

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}\label{fig:map}Location of the three
coastal communities studied (A). Isla Natividad (B) is located off the
Baja California Peninsula, Maria Elena (C) and Punta Herrero (D) are
located in the Yucatan Peninsula. Blue polygons represent the TURFs, and
red polygons the marine reserves.}
\end{figure}

@@

Socioeconomic data come from landing receipts reported to the National
Commission for Aquaculture and Fisheries (\emph{Comisión Nacional de
Acuacultura y Pesca}; CONAPESCA). Data contain monthly lobster landings
(Kg) and value (MXP) from 2000 to 2014 for cooperatives with and without
marine reserves (\textbf{Fig S1}). This information was aggregated by
year, and economic values were adjusted by the Consumer Price Index
\citep{oecd_2017-VV} as:

@@

Data for the qualitative analysis of the social-ecological system were
collected at the community--level from official documents used in the
creation and designation of the marine reserves
\citep{dof_website_2012,dof_website_2013} and based on the authors'
experience and knowledge of the communities. These include information
on the resource system, the resource units, actors, and the governance
system itself (\textbf{S1 Table}).

@@

We evaluate the effect that marine reserves have had on four biological
and two socioeconomic indicators (Table \ref{table:indicators}). Recall
that reserves were implemented to protect lobster and other benthic
invertebrates. However, we also use the available fish data, since
finfish are targeted in the off-season and because there might be
associated cobenefits of the reserve.

@@

\caption{\label{tab:unnamed-chunk-3}\label{table:indicators}List of indicators used to evaluate the effectiveness of marine reserves, grouped by category.}
\centering
\begin{tabular}[t]{l|l|l}
\hline
Category & Indicador & Units\\
\hline
Biological & Lobster density & org / $m^2$\\
\hline
Biological & Invertebrate density & org / $m^2$\\
\hline
Biological & Fish biomass & Kg / $m^2$\\
\hline
Biological & Fish density & org / $m^2$\\
\hline
Socioeconomic & Income from target species & M MXP\\
\hline
Socioeconomic & Landings from target species & Metric Tonnes\\
\hline
\end{tabular}
\end{table}

We use a difference-in-differences analysis to evaluate the biological
indicators. This approach allows us to estimate the effect that the
reserve has on the biological indicators by comparing trends across time
and treatments (\emph{i.e.} reserve / control sites
\citet{moland_2013-VP,Villasenor-Derbez_2018}). The analysis is
performed with a multiple linear regression of the form:

@@

Where year-fixed effects are represented by \(\gamma_{it} Year_t\), and
\(\beta Zone_i\) captures the difference between reserve (\(Zone = 1\))
and control (\(Zone = 0\)) sites. The interaction term
\(\lambda_{it} Year_t\times Zone_i\) represents the mean change in the
indicator inside the reserve, for year \(t\), with respect to the year
of implementation in the control site (See Table \ref{table:com_sum}).
When evaluating biomass and densities of the entire benthic or fsh
communities, we include \(\sigma_j\) to control for species-fixed
effects.

@@

\begin{equation}
I = \alpha + \gamma_{t} Year_t + \beta Treated_i + \lambda_{t} Year_t\times Treated_i + \sigma_jCom_j +\epsilon
\label{eqn:soc_reg}
\end{equation}

The model interpretation remains as for Eq \ref{eqn:reg_bio}, but in
this case the \(Treated\) dummy variable indicates if the community has
a reserve (\(Treated = 1\)) or not (\(Treated = 0\)) and \(\sigma_jCom\)
captures community-level fixed-effects. These regressions allows us to
make a causal link between the implementation of marine reserves and the
observed trends by accounting for temporal and spatial dynamics
\citep{depalma_2018}. The effect of the reserve is captured by the
\(\lambda_t\) coefficient, and represents the difference observed
between the control site before the implementation of the reserve and
the treated sites at time \(t\) after controlling for other time and
space variations (\emph{i.e.} \(\gamma_t\) and \(\beta\) respectively).
All model coefficients were estimated via ordinary least-squares and
heteroskedastic-robust standard errors \citep{zeileis_2004-7n}. All
analyses were performed in R 3.5.0 and R Studio 1.1.453 \citep(R\_2018).

@@

The following sections present the effect that marine reserves had on
each of the biological and socioeconomic indicators for each coastal
community. Results are presented in terms of the difference through time
and across sites, relative to the control site on the year of
implementation (\emph{i.e.} effect size \(\lambda_t\)). We also provide
an overview of the governance settings of each community, and discuss
how these dimensions might be related to the effectiveness of the
reserves.

@@

Indicators showed idiosyncratic responses through time for each
community. Figure \ref{fig:indicators}A shows positive effect sizes for
lobster densities for Isla Natividad and Punta Herrero during the first
years, but the effect is eroded through time. These effects are in the
order of 0.2 extra organisms \(m^{-2}\) but are not significantly
different from zero (\(p > 0.05\)). Lobster densities were only
significantly positive for Isla Natividad on the sixth year (\emph{i.e.}
2012; \(p < 0.05\)), a year after the hypoxia events described by
\citet{micheli_2012-EU} caused mass mortality of organisms. Likewise, no
changes were detected in fish biomass or invertebrate and fish densities
(\ref{fig:indicators}B-D), where effect sizes oscillated around zero
without clear trends. Full tables with model coefficients are presented
in the supplementary materials (\textbf{S2 Table}, \textbf{S3 Table},
\textbf{S4 Table}).

@@

Lobster landings and revenue were only available for Isla Natividad and
Maria Elena (Fig \ref{fig:lobsters}). For all years befor
implementation, the effect sizes are close to zero, indicating that the
control and treatment sites track each other well and that these are
plausible controls. However, effect sizes do not change after the
implementation of the reserve. Again, the negative coeficient observed
for Isla Natividad on year 5 correspond to the 2011 hypoxia events. The
only positive change observed in lobster landings is for Isla Natividad
in 2014 (\(p < 0.1\)). The trhee years of post-implementation data for
Maria Elena do not show a significant effect of the reserve. Isla
Natividad shows higher revenues after the iumplementation of the
reserve, as compared to the control communities. However, these changes
are also not ignificantly different and present an increased variation.
All regression coefficients for each community and indicator are
presented in \textbf{S5 Table}.

@@

Although we have little information on the social dimension of these
fisheries, we can use the social-ecologycal systems framework
(\textbf{S1 Table}) to analyze the performance of each governance system
(\textbf{S6 Table}).

@@

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}\label{fig:indicators}Effect sizes for
marine reserves from Isla Natividad (IN; red cirlcles), Maria Elena (ME;
blue triangles), and Punta Herrero (PH; green squares) for lobster
densities (\emph{Panulirus spp}; A), fish biomass (B), invertebrate
densities (C), and fish densities (D). Plots are ordered by survey type
(left column: invertebrates; right column: fish). Points are jittered
hotizontally to avoid overplotting. Points indicate the effect size, and
errorbars standard errors. Years have been centered to year of
implementation.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Villasenor-Derbez_files/figure-latex/unnamed-chunk-6-1.pdf}
\caption{\label{fig:unnamed-chunk-6}\label{fig:lobsters}Effect sizes for
lobster catches (A) and revenues (B) in at Isla Natividad (IN; red
circles) and Maria Elena (ME; blue triangles)}
\end{figure}

@@

Map of control and treated sites in A and control and treated landings
in B

@@

\paragraph*{S4 Figure}
\label{S3_Figure}
